Effects of prenatal CORT and temperature on the response to a stressor

### 2025-07-09 ms.qmd
updates

---
title: "Effects of prenatal CORT and temperature on the response to a stressor"
docx: 
  number_sections: true
  linestretch: 2
  fontsize: 12pt
  mainfont: Times New Roman
bibliography: "./bib/refs.bib"
csl: "./bib/animal-behaviour.csl"
reference-doc: "./bib/tmpl.docx"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig.    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

Animals constantly face events that can disrupt their homeostasis. These challenges can be acute, such as a sudden predator attack, or long-term, like prolonged droughts or habitat degradation [@sapolsky_how_2000; @mcewen_concept_2003]. Regardless of the nature of the challenge, organisms must adjust to these alterations to maintain their physiological and behavioural integrity [@wingfield2002endocrine]. In vertebrates, the ability to cope with allostasis is mediated by the hypothalamic-pituitary-adrenal/interrenal (HPA/HPI) axis [@sapolsky_how_2000]. This system elicits a physiological response that results in the secretion of glucocorticoids (GCs) [@sapolsky_how_2000]. Glucocorticoids are a group of hormones that help reallocate energetic resources to restore homeostasis following disruption [@sapolsky_how_2000; @mcewen_concept_2003]. For example, when an animal faces a predatory attack, GCs mobilise energy to support increased locomotor activity, vigilance, or attention to avoid the threat [@trompeter2011invader].

In the short term, GCs help organisms cope with acute challenges and return to homeostasis. However, long-term exposure to GCs can carry significant metabolic costs [@picard_mitochondria_2014]. Prolonged elevations in GCs can lead to immune suppression, impaired growth, or permanent neural alterations [@mcewen2017neurobiological]. From a behavioural perspective, reacting to stressors can reduce the time and energy available for essential activities such as foraging, thermoregulation, or social interactions [@belliure2004behavioral; @martin2024blind]. As such, animals are predicted to evolve mechanisms to appropriately respond to acute stressors while avoiding the costs in the long-term. For example, by regulating their responses to homotypic (repeated and similar) stressors, particularly those that are predictable and non-lethal [@grissom2009habituation]. Habituation - the reduction of physiological responses elicited by exposure to a homotypic stressor [@pfister1979glucocorticosterone] - can be crucial to avoid the long-term costs of chronic stress without compromising the response to acute stressors.

Importantly, the process of physiological habituation shares key functional characteristics with the cognitive process of habituation - a form of non-associative learning where individuals reduce their responsiveness to a repeated stimulus [@thompson1966habituation]. In both cases, organisms are repeatedly exposed to a stimulus/homotypic stressor and progressively decrease their response towards it [@grissom2009habituation]. In fact, both behavioural and physiological habituation often co-occur. For example, rhesus macaques (_Macaca mulatta_) showed a decline in blood cortisol concentrations across consecutive days of restraint together with decreased vocalizations and behavioral agitation [@ruys2004behavioral]. How both types of habituation interplay with each other depends on the nature of the stressor. Physiological habituation does not occur when animals are faced with systemic stressors like hypoglycemia or ether exposure whereas habituation to processive stressors — those involving primarily psychological qualities, such as restraint or simulated predation - is frequent [@grissom2009habituation]. However, physiological habituation alone is insufficient to explain the behavioural patterns observed when animals are repeteadly exposed to a homotypic processive stressor [@jaferi2006corticosterone; @grissom2009habituation]. Adrenalectomised rats - incapable of physiological habituation - still exhibit cognitive habituation [@jaferi2006corticosterone], suggesting a degree of independence between the two processes. While the HPA/HPI axis is paramount for physiological habituation, cognitive habituation also involve neural mechanisms related to sensory processing, learning, and behavioural control such as the limbic system, the superior colliculus, or the optic tectum [@grissom2009habituation; @dutta2014saliency]. Habituation to processive stressors is therefore likely to involve complex interactions between physiological and cognitive processes.

Crucially, the effectiveness of either system depends on how well the underlying neural structures develop and integrate. Many of these brain regions mature concurrently and can be highly plastic during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017]. Early-life conditions can shape the structure and function of both neural systems, with long-term consequences for how individuals respond to acute and chronic stressors [@vargas2016early; @van2017modulation; @zito2017early; @crino_under_2020; @crino2024eggs]. One of the key mediators of early-life effects is GC exposure. Elevations in GCs during development exert potent effects on HPA axis function, altering the response to stressors later in life [@costantini2011meta; @vargas2016early; @zito2017early; @crino2024eggs]. For example, zebra finches (_Taeniopygia guttata_) nutritionally stressed during early-life had higher levels of CORT than controls at post hatching day 25, while periodic maternal separation increased basal corticosterone levels in rats (_Rattus norvegicus_) [@vargas2016early; @zito2017early]. In parallel, developmental GC exposure can impact brain regions involved in learning, memory, and behavioural flexibility, affecting cognitive processes relevant to cognitive habituation [@lemaire_prenatal_2000; @zhu_prenatal_2004; @costantini2011meta]. Therefore, early-life exposure to GCs can have long-lasting effects on the response to both acute and chronic stressors, potentially leading to maladaptive responses in adulthood.

Other environmental factors may also shape stress-related phenotypes, either independently or in combination with GCs. In ectotherms, early thermal conditions play a critical role in cognitive development [@amiel_effects_2017; @abayarathna_effects_2020]. Incubation temperature affects cognitive abilities such as learning, memory, or perception, likely through changes in brain function and structure [@amiel_smart_2011; @amiel_hotter_2012; @amiel_effects_2017; @dayananda_incubation_2017; @abayarathna_effects_2020]. In addition, early thermal environment can also affect the development of the HPA/HPI axis [@crino_under_2020; @crino2024eggs]. For example, in the delicata skink (_Lampropholis delicata_) individuals incubated at cooler temperatures showed increased baseline CORT concentration in the blood [@crino2024eggs]. Although these differences are believed to be driven by size rather than direct endocrine effects [@crino2024eggs], they can still have important long-term consequences for how individuals respond to acute and chronic stressors. Importantly, incubation temperature and prenatal CORT may not act independently [see @wingfield_comparative_2008; @Crino_2023], and their interaction could produce complex, trait-specific outcomes that shape both physiological and behavioural responses to stress. Yet, the interactive effects of these early-life conditions on stress responses remain largely unexplored.

@@

We used two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). These are oviparous, generalist, small (∼35-55 mm snout-vent length (SVL)) lizards that are sympatric in suburban areas throughout south-eastern Australia [@chapple_know_2011; @chapple_biology_2014]. However, _L. delicata_ has been reported as a successful invader in New Zealand and Hawaii, while there are not such reports for _L. guichenoti_ [@chapple_know_2011; @chapple_biology_2014]. Invassivenes does not seem to be related to differences in cognitive abilities [@bezzina2014does; @recio2025early; @recio2025early], but seems to be associated with differences in personality [@chapple_know_2011], which may be linked to differences in their stress response [@koolhaas1999coping].

### Subjects treatments, collection, and housing

*Breeding colony* - Lizards came from a breeding colony established in the laboratory since 2019. This colony consisted of 270 adults of _L. delicata_ and 180 adults of _L. guichenoti_ housed in plastic containers (41.5 L x 30.5 W x 21 H cm) with two males and four females per enclosure. Enclosures were provided with shelter, nonstick matting, and several small water dishes. The lizards were given water daily and were fed approx. 40 mid-size crickets (_Acheta domestica_) per enclosure three days a week. The crickets were dusted with calcium weekly and multivitamin and calcium biweekly. Room temperature was set to 22-24 ºC, but we also provided the enclosures with a heat chord and a heat lamp following a 12 h light:12 h dark cycle keeping warm side of enclosures at 34 ºC.

*Eggs collection and incubation* - Eggs were collected between mid-October 2022 to the end of February 2023. We placed a small box (12.5 L x 8.3 W x 5 H cm) with moist vermiculite in one side of the communal enclosures to provide females with a place to lay the eggs. These boxes were checked three days a week. After egg collection, we measured length and width with a digital caliper to the nearest 0.1 mm and weighted the eggs with a (OHAUS, Model spx123) digital scale with an accuracy of ± 0.001 g error. Then eggs were treated with CORT or vehicle (see CORT and temperature manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in two incubators at two different temperatures (see CORT and temperature manipulation below) until hatching.

@@

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them at cold (23 ± 3 ºC) or hot (28 ± 3 ºC) conditions (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or an equal volume of 100% Ethanol (Control treatment). We selected doses based on our previous study where CORT treatment increased mean yolk CORT levels by approximately 2 standard deviations above the mean natural concentration [@crino2024eggs]. Mean incubation temperatures represent the lower and upper limits of the natural range of nest temperatures in _L. delicata_ [mean nesting temperatures = 27.4 ºC; @cheetham2011embryonic]. Nonetheless, the higher temperature treatment (28 ºC) was above the thermal optima estimated for _L. delicata_ [T~opt~ = 25 ºC; @pettersen2023maternal] and reached more stressful temperatures daily as fluctuations occurred (i.e., 31 ºC).

@@

Both before and after the completion of the behavioural tests, lizard mass was measured to the nearest 0.001 g using a digital scale. Change in mass was calculated as the difference between final minus the initial mass and then rescaled by adding the smallest mass change observed in the dataset, to avoid negative values. This rescaled change in mass was used as a proxy for the physiological cost of repeated stress exposure. To control for hunger levels and potential effects of food ingestion on mass change, we also recorded the number of crickets ingested by each lizard during the acclimation period. To do so, we placed a known number of crickets in each enclosure after each of the trials, and counted the number of crickets remaining after 24 hours. The total number of crickets ingested during the eight days of tests was then used as a covariate in the analyses of mass change. The experimental design is summarised in @fig-Methods B.

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Scheme of our experimental design. In panel A, the early environment manipulations resulting in four experimental conditions. In panel B, the timeline of the postnatal stress-response assay. On Day 0, lizard mass was recorded. From Days 1–8, lizards were subjected daily to a simulated predator attack (chased with a soft paintbrush for 1 minute), followed by a 1-hour behavioural recording. Variables measured included freezing duration (latency to move), latency to shelter, and emergence from shelter. Final mass was measured on Day 9."

@@

We used as the response variables the latency to move, latency to shelter, and emergence from the shelter after the simulated attack, as well as the change in mass after the eight days of stress exposure. We fitted separate models for each response variable. For the three behavioural variables, we included the prenatal condition (Control-Cold, CORT-Cold, Control-Hot, CORT-Hot), the day of the trial, and their interaction as fixed effects. To facilitate interpretation of the model intercepts, trial day was transformed so that day 1 in the experiment corresponds to day 0 in the model (i.e. the intercept reflects lizard behaviour in response to the first exposure to the stressor). For the variable mass, we included the prenatal condition, the total number of crickets ingested during the eight days of stress exposure, and their interaction as predictors.

Lizard identity was included as a random factor in the behavioural models, but it was not included when mass was the response. For all models, we included the clutch as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple2015deliinvLHI]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. Additionally, previous research has shown that clutches are generally sired by a single male, but sperm storage can occur [@kar2023heritability]. Given our partial split-clutch design, and the fact that maternal effects are expected to be stronger than paternal effects in these species, including the clutch as a random factor should account for the effects of parental identity.

We used the posterior distributions of model parameters to test for differences between prenatal conditions. For behavioural variables, differences in the model intercepts between conditions were interpreted as differences in the acute response to a stressor (i.e. the first day of exposure), while differences in the slopes were interpreted as differences in behavioural adjustment across repeated exposures (i.e. chronic stress response). For mass change, between treatments difference reflect differences in energetic cost of the chronic stressor. We used the 95% Highest Posterior Density Intervals (95% HPDI) using the hdi function in bayestestR [@bayestestR] to test if the contrasts between treatments or the slopes for each treatment were different from zero.

@@

_Lampropholis delicata_: Incubation at cold temperatures decreased the latency to move after a simulated attack on the first day (Intercept (denoted hereafter as I)~Hot~ - I ~Cold~ = `r deli_move_contrasts$median_Temperature_int`; 95% Highest Posterior Density Intervals (95% HPDI) = [`r deli_move_contrasts$X2.5.HDI_Temperature_int`, `r deli_move_contrasts$X97.5.HDI_Temperature_int`]), while neither CORT or the interaction have any effect on latency to shelter or emergence (see @fig-results_beh_deli A, D, G and Table S1 in _Supplementary Material_). We did not find any effects of prenatal temperature, CORT or their interaction on latency to shelter or emergence probability after an acute stressor (see @fig-results_beh_deli A, D, G and Table S1 in _Supplementary Material_).

@@

## Discussion

>>>>

---
title: "Effects of prenatal CORT and temperature on the response to a stressor"
format:
  docx: 
    reference-doc: "./bib/tmpl.docx"
bibliography: "./bib/refs.bib"
csl: "./bib/animal-behaviour.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig.    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

Animals constantly face events that can disrupt their homeostasis. These disruptions can be acute, such as a sudden predator attack, or long-term, like prolonged droughts or habitat degradation [@sapolsky_how_2000; @mcewen_concept_2003]. Regardless of the nature of the challenge, organisms must adjust to allostasis to maintain their physiological and behavioural integrity [@wingfield2002endocrine]. In vertebrates, the ability to cope with disruptions in homeostasis is carried out by the hypothalamic-pituitary-adrenal/interrenal (HPA/HPI) axis [@sapolsky_how_2000]. This system elicits a physiological response, also known as the stress response, mediated by glucocorticoids (GCs), a group of hormones that facilitate the reallocation of energetic resources to restore homeostasis [@sapolsky_how_2000; @mcewen_concept_2003]. For example, when an animal faces a predatory attack, GCs mobilise energy to support increased locomotor activity, vigilance, or attention to avoid the threat [@trompeter2011invader]. In the short term, GCs help organisms cope with acute challenges. However, long-term exposure to GCs can carry significant metabolic costs [@picard_mitochondria_2014]. Prolonged elevations in GCs can lead to immune suppression, impaired growth, or permanent neural alterations [@mcewen2017neurobiological]. Furthermore, reacting to stressors can reduce the time and energy available for essential activities such as foraging, thermoregulation, or social interactions [@belliure2004behavioral; @martin2024blind]. As such, animals are predicted to evolve mechanisms to appropriately respond to acute stressors while avoiding the costs in the long term. For example, by regulating their responses to homotypic (repeated and similar) stressors, particularly those that are predictable and non-lethal [@grissom2009habituation]. Habituation - the reduction of physiological responses elicited by exposure to a homotypic stressor [@pfister1979glucocorticosterone] - can be crucial to avoid the long-term costs of chronic stress without compromising the response to acute stressors.

Importantly, the process of physiological habituation shares key functional characteristics with the cognitive process of habituation - a form of non-associative learning where individuals reduce their responsiveness to a repeated stimulus [@thompson1966habituation]. In both cases, organisms are repeatedly exposed to a stimulus/homotypic stressor and progressively decrease their response towards it [@grissom2009habituation]. In fact, both behavioural and physiological habituation often co-occur. For example, rhesus macaques (_Macaca mulatta_) showed a decline in blood cortisol concentrations across consecutive days of restraint together with decreased vocalizations and behavioral agitation [@ruys2004behavioral]. How both types of habituation interplay with each other depends on the nature of the stressor. Physiological habituation does not occur when animals are faced with systemic stressors like hypoglycemia or ether exposure, whereas habituation to processive stressors — those involving primarily psychological qualities, such as restraint or simulated predation - is frequent [@grissom2009habituation]. However, physiological habituation alone is insufficient to explain the behavioural patterns observed when animals are repeatedly exposed to a homotypic processive stressor [@jaferi2006corticosterone; @grissom2009habituation]. Adrenalectomised rats - incapable of physiological habituation - still exhibit cognitive habituation [@jaferi2006corticosterone], suggesting a degree of independence between the two processes. While the HPA/HPI axis is paramount for physiological habituation, cognitive habituation also involves neural mechanisms related to sensory processing, learning, and behavioural control, such as the limbic system, the superior colliculus, or the optic tectum [@grissom2009habituation; @dutta2014saliency]. Habituation to processive stressors is therefore likely to involve complex interactions between physiological and cognitive processes.

Crucially, the effectiveness of either system depends on how well the underlying neural structures develop and integrate. Many of these brain regions mature concurrently and can be highly plastic during early stages of development [@coomber_independent_1997; @zhu_prenatal_2004; @amiel_effects_2017]. Early-life conditions can shape the structure and function of both neural systems, with long-term consequences for how individuals respond to acute and chronic stressors [@vargas2016early; @van2017modulation; @zito2017early; @crino_under_2020; @crino2024eggs]. One of the key mediators of early-life effects is GC exposure. Elevations in GCs during development exert potent effects on HPA axis function, altering the response to stressors later in life [@costantini2011meta; @vargas2016early; @zito2017early; @crino2024eggs]. For example, zebra finches (_Taeniopygia guttata_) nutritionally stressed during early-life had higher levels of CORT than controls at post-hatching day 25, while periodic maternal separation increased basal corticosterone levels in rats (_Rattus norvegicus_) [@vargas2016early; @zito2017early]. In parallel, developmental GC exposure can impact brain regions involved in learning, memory, and behavioural flexibility, affecting cognitive processes relevant to cognitive habituation [@lemaire_prenatal_2000; @zhu_prenatal_2004; @costantini2011meta]. Therefore, early-life exposure to GCs can have long-lasting effects on the response to both acute and chronic stressors, potentially leading to maladaptive responses in adulthood.

Other environmental factors may also shape stress-related phenotypes, either independently or in combination with GCs. In ectotherms, early thermal conditions play a critical role in cognitive development [@amiel_effects_2017; @abayarathna_effects_2020]. Incubation temperature affects cognitive abilities such as learning, memory, or perception, likely through changes in brain function and structure [@amiel_smart_2011; @amiel_hotter_2012; @amiel_effects_2017; @dayananda_incubation_2017; @abayarathna_effects_2020]. In addition, early thermal environment can also affect the development of the HPA/HPI axis [@crino_under_2020; @crino2024eggs]. For example, in the delicata skink (_Lampropholis delicata_), individuals incubated at cooler temperatures showed increased baseline CORT concentration in the blood [@crino2024eggs]. Although these differences are believed to be driven by size rather than direct endocrine effects [@crino2024eggs], they can still have important long-term consequences for how individuals respond to acute and chronic stressors. Importantly, incubation temperature and prenatal CORT may not act independently [see @wingfield_comparative_2008; @Crino_2023], and their interaction could produce complex, trait-specific outcomes that shape both physiological and behavioural responses to stress. Yet, the interactive effects of these early-life conditions on stress responses remain largely unexplored.

@@

We used two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). These are oviparous, generalist, small (∼35-55 mm snout-vent length (SVL)) lizards that are sympatric in suburban areas throughout south-eastern Australia [@chapple_know_2011; @chapple_biology_2014]. However, _L. delicata_ has been reported as a successful invader in New Zealand and Hawaii, while there are no such reports for _L. guichenoti_ [@chapple_know_2011; @chapple_biology_2014]. Invassiveness does not seem to be related to differences in cognitive abilities [@bezzina2014does; @recio2025early; @recio2025early], but seems to be associated with differences in personality [@chapple_know_2011], which may be linked to differences in their stress response [@koolhaas1999coping].

### Collection and housing

*Breeding colony* - Lizards came from a breeding colony established in the laboratory since 2019. This colony consisted of 270 adults of _L. delicata_ and 180 adults of _L. guichenoti_ housed in plastic containers (41.5 L x 30.5 W x 21 H cm) with two males and four females per enclosure. Enclosures were provided with shelter, nonstick matting, and several small water dishes. The lizards were given water daily and were fed approx. 40 mid-size crickets (_Acheta domestica_) per enclosure three days a week. The crickets were dusted with calcium weekly and multivitamins and calcium biweekly. Room temperature was set to 22-24 ºC, but we also provided the enclosures with a heat cord and a heat lamp following a 12 h light:12 h dark cycle, keeping the warm side of the enclosures at 34 ºC.

*Eggs collection and incubation* - Eggs were collected between mid-October 2022 to the end of February 2023. We placed a small box (12.5 L x 8.3 W x 5 H cm) with moist vermiculite on one side of the communal enclosures to provide females with a place to lay the eggs. These boxes were checked three days a week. After egg collection, we measured length and width with a digital caliper to the nearest 0.1 mm and weighed the eggs with a (OHAUS, Model spx123) digital scale with an accuracy of ± 0.001 g error. Then eggs were treated with CORT or vehicle (see CORT and temperature manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in two incubators at two different temperatures (see CORT and temperature manipulation below) until hatching.

@@

To test the interactive effects of CORT and incubation temperature, we manipulated CORT concentrations in eggs and incubated them at cold (23 ± 3 ºC) or hot (28 ± 3 ºC) conditions (see @fig-Methods A). We used a partial split clutch design where eggs from a given clutch were distributed equally across the four treatments when clutch sizes were larger than four and randomly across treatments when less than four. Eggs were topically supplied with either 5 µL of crystalline corticosterone (Sigma, Cat. No. C2505) dissolved in 100% ethanol at a final 10 pg CORT/mL concentration (CORT treatment), or an equal volume of 100% Ethanol (Control treatment). We selected doses based on our previous study, where CORT treatment increased mean yolk CORT levels by approximately 2 standard deviations above the mean natural concentration [@crino2024eggs]. Mean incubation temperatures represent the lower and upper limits of the natural range of nest temperatures in _L. delicata_ [mean nesting temperatures = 27.4 ºC; @cheetham2011embryonic]. Nonetheless, the higher temperature treatment (28 ºC) was above the thermal optima estimated for _L. delicata_ [T~opt~ = 25 ºC; @pettersen2023maternal] and reached more stressful temperatures daily as fluctuations occurred (i.e., 31 ºC).

@@

Both before and after the completion of the behavioural tests, lizard mass was measured to the nearest 0.001 g using a digital scale. Change in mass was calculated as the difference between the final minus the initial mass and then rescaled by adding the smallest mass change observed in the dataset, to avoid negative values. This rescaled change in mass was used as a proxy for the physiological cost of repeated stress exposure. To control for hunger levels and potential effects of food ingestion on mass change, we also recorded the number of crickets ingested by each lizard during the acclimation period. To do so, we placed a known number of crickets in each enclosure after each of the trials, and counted the number of crickets remaining after 24 hours. The total number of crickets ingested during the eight days of tests was then used as a covariate in the analyses of mass change. The experimental design is summarised in @fig-Methods B.

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Scheme of our experimental design. In panel A, the early environment manipulations leading to four experimental conditions. In panel B, the timeline of the postnatal stress-response assay. On Day 0, lizard mass was recorded. From Days 1–8, lizards were subjected daily to a simulated predator attack (chased with a soft paintbrush for 1 minute), followed by a 1-hour behavioural recording. Variables measured the latency to move, the latency to shelter, and the probability of emergening from the shelter in 40 minutes after the stressor. Final mass was measured on Day 9."

@@

We used as the response variables the latency to move, the latency to shelter, and the emergence probability, as well as the change in mass after the eight days of stress exposure. We fitted separate models for each response variable. For the three behavioural variables, we included the prenatal condition (Control-Cold, CORT-Cold, Control-Hot, CORT-Hot), the day of the trial, and their interaction as fixed effects. To facilitate interpretation of the model intercepts, trial day was transformed so that day 1 in the experiment corresponds to day 0 in the model (i.e., the intercept reflects lizard behaviour in response to the first exposure to the stressor). For the variable mass, we included the prenatal condition, the total number of crickets ingested during the eight days of stress exposure, and their interaction as predictors.

Lizard identity was included as a random factor in the behavioural models, but it was not included when mass was the response. For all models, we included the clutch as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple2015deliinvLHI]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. Additionally, previous research has shown that clutches are generally sired by a single male, but sperm storage can occur [@kar2023heritability]. Given our partial split-clutch design and the fact that maternal effects are expected to be stronger than paternal effects in these species, including the clutch as a random factor should account for the effects of parental identity.

We used the posterior distributions of model parameters to test for differences between prenatal conditions. For behavioural variables, differences in the model intercepts between conditions were interpreted as differences in the acute response to a stressor (i.e., the first day of exposure), while differences in the slopes were interpreted as differences in behavioural adjustment across repeated exposures (i.e., chronic stress response). For mass change, between-treatments differences reflect differences in the energetic cost of the chronic stressor. We used the 95% Highest Posterior Density Intervals (95% HPDI) using the hdi function in bayestestR [@bayestestR] to test if the contrasts between treatments or the slopes for each treatment were different from zero.

@@

_Lampropholis delicata_: Incubation at cold temperatures decreased the latency to move after a simulated attack on the first day (Intercept (denoted hereafter as I)~Hot~ - I ~Cold~ = `r format_dec(exp(deli_move_contrasts$median_Temperature_int), 3)`; 95% Highest Posterior Density Intervals (95% HPDI) = [`r deli_move_contrasts$X2.5.HDI_Temperature_int`, `r deli_move_contrasts$X97.5.HDI_Temperature_int`]), while neither CORT or the interaction have any effect on latency to shelter or emergence (see @fig-results_beh_deli A, D, G and Table S1 in _Supplementary Material_). We did not find any effects of prenatal temperature, CORT or their interaction on latency to shelter or emergence probability after an acute stressor (see @fig-results_beh_deli A, D, G and Table S1 in _Supplementary Material_).

@@

## Discussion    
Our study shows that early-life conditions can influence how skinks respond to a processive stressor, but the effects differ between species, the nature of th stressor, and the behaviour evaluated. _Lampropholis delicata_ incubated at warmer temperatures were less reactive to a simulated predator attack in the first trial, suggesting that early thermal conditions can shape the behavioural response to an acute stressor. In contrast, _L. guichenoti_ showed no treatment effects. Habituation patterns also differed. _Lampropholis delicata_ increased the probability of emerging from the shelter, suggesting some habituation to the stressor. However, none of the treatments showed habituation in latency to move or go to shelter. No behavioural change was detected in _L. guichenoti_ either. Finally, we found no evidence of physiological costs: individuals generally maintained or gained body mass, regardless of treatment. Together, these findings suggest that incubation temperature and prenatal hormone exposure shape stress-related behaviour in a species-specific manner, potentially reflecting differences in flexibility or stress physiology.

#### Cold incubation temperature increases the response to an acute stressor in _L. delicata_ but not in _L. guichenoti_          
Our results demonstrate that incubation temperature modulates the behavioural response to an acute stressor in _L. delicata_ but not in _L. guichenoti_. Hot-incubated _L. delicata_ spent less time immobile following a simulated predatory attack, suggesting a faster recovery from acute stress. **This result is consistent with the idea that shorter immobility duration (i.e., shorter freezing time) reflects a less reactive behavioural response, often interpreted as reduced perceived risk or more efficient coping [REFs].** Thus, the reduced latency to move of hot-incubated individuals may indicate reduced stress reactivity. In _L. delicata_, individuals incubated at 23 ºC exhibit elevated baseline CORT [@crino2024eggs], potentially enhancing sensitivity to stress. Therefore, the heightened acute response in cold-incubated individuals may be explained by elevated endogenous CORT levels. For example, **E.G. about CORT and response**.

In contrast, early thermal environment had no effect on _L. guichenoti_'s response to an acute stressor. The species-specific response also warrants discussion. L. delicata exhibited treatment-dependent variation in response to acute stress, while L. guichenoti did not. This difference may reflect distinct developmental plasticity or stress-coping strategies between species. Given that L. delicata is a successful invasive species, greater sensitivity to early-life thermal environments and a more flexible behavioural response to acute threats may confer adaptive advantages in novel environments [REFs]. In contrast, the more canalised stress response in L. guichenoti could reflect reduced behavioural plasticity, potentially limiting its adaptability in changing environments.

Strikingly, we found no effects of prenatal CORT treatment on the acute stress response in either species. Glucocorticoid elevation during development are known to modulate stress-related phenotypes, including behavioural responsiveness to processive stressors [**REF...before?? I might have another in the intro**]. Previous studies have shown that

#### Habituation to stressors varies between species and treatments
Across repeated exposures to the simulated predatory attack, only L. delicata individuals from the CORT-Cold treatment increased their latency to move. Rather than indicating habituation, this pattern may reflect accumulating allostatic load. Habituation is typically inferred from a reduction in stress-related behaviours over time (e.g., decreased latency to resume activity), whereas an increase in latency could reflect mounting stress or impaired coping [@grissom2009habituation]. Thus, the behavioural trajectory of the CORT-Cold group may suggest that these animals became less, rather than more, tolerant to the repeated stressor.

Why did this pattern emerge only in the CORT-Cold L. delicata group? While our CORT treatment did not significantly elevate baseline hormone levels across groups, previous findings indicate that cold incubation alone raises baseline CORT [@crino2024eggs]. The combination of prenatal CORT and cold temperature may therefore have produced a cumulative or synergistic effect on HPI axis development and reactivity. This effect might not have been detectable in our acute response measures, but could become apparent over repeated exposures as stress accumulates.

Alternatively, cognitive explanations are also plausible. The cold-CORT combination may impair neural development in regions critical for behavioural flexibility and non-associative learning. Although our prior work found no significant impact of this combination on associative learning (e.g., colour discrimination and reversal learning; @recio2025cognitive; @recio2025early), cognitive processes underpinning habituation may rely on different neural substrates. The hippocampus and prefrontal cortex, implicated in behavioural inhibition and decision-making, may be differentially affected by early-life CORT and temperature exposure. Region-specific developmental trajectories could account for divergent effects across learning types.

In contrast, L. guichenoti showed no evidence of behavioural change across repeated stress exposures, regardless of treatment. This further supports the notion of reduced behavioural flexibility or a more rigid stress-response phenotype in this species. It is noteworthy that L. delicata displayed treatment-specific changes even when these did not conform to classical habituation. Such responsiveness may reflect a greater capacity to modulate behaviour based on environmental cues, a trait often associated with invasive potential [REFs].

Interestingly, across all treatments, L. delicata increased their probability of emerging from shelter within 40 minutes of the stressor as the experiment progressed, suggesting that overall risk perception decreased over time. This pattern is consistent with habituation at the level of decision-making rather than immediate post-threat responses (i.e., latency to move). L. guichenoti, in contrast, exhibited no such trend. This species difference in shelter-emergence behaviour provides further support for differential flexibility in stress evaluation and risk assessment. The divergence may again be rooted in differences in developmental plasticity or evolutionary history, with L. delicata potentially possessing a more labile stress-response system suited to fluctuating environments.

The absence of time effects in latency to move or go to shelter for the other treatments in L. delicata suggests that habituation was limited and potentially context- or trait-dependent. Taken together, our findings highlight the complexity of stress habituation and the need to disentangle multiple behavioural and physiological components to fully understand coping strategies.

#### Physiological costs of the stressor were low and did not differ between treatments or species
Despite repeated exposure to a stressor over several days, we found no clear evidence of physiological costs in either species. Most individuals gained mass over the course of the experiment, and there were no treatment or species differences in mass change. This suggests that the imposed stressor, while sufficient to elicit behavioural responses, did not impair growth or energy balance under our conditions. This result is somewhat surprising, given the literature linking chronic stress to impaired growth, particularly in juvenile ectotherms [REFs].

One possible explanation is that the stressor we employed, while ecologically relevant, was not sufficiently intense or frequent to produce measurable physiological effects. In natural settings, stressors may be more variable and compound with other challenges (e.g., food scarcity, thermal extremes), resulting in more pronounced physiological consequences. Our controlled environment, consistent food availability, and short stress exposure duration may have buffered such effects.

Another consideration is that mass change may not fully capture the physiological costs of stress. Other endpoints, such as oxidative stress markers, telomere attrition, or immune function, may be more sensitive to sublethal chronic stress effects [REFs]. Future studies incorporating these physiological markers would provide a more comprehensive picture of the costs associated with repeated stress.

Finally, it is important to note that our behavioural findings suggest differences in stress evaluation and habituation across treatments and species, even in the absence of overt physiological changes. This dissociation underscores the importance of measuring multiple dimensions of the stress response and not relying solely on growth or condition as proxies for stress burden.

#### Conclusion
Our study demonstrates that early-life environments—specifically incubation temperature and prenatal glucocorticoid exposure—shape behavioural responses to acute and repeated stressors in skinks. These effects were species-specific, with L. delicata showing greater responsiveness to early-life conditions than L. guichenoti. Cold incubation increased acute stress reactivity in L. delicata, and the combination of cold and prenatal CORT impaired habituation, potentially via elevated baseline CORT or neural alterations. In contrast, L. guichenoti appeared less flexible in both acute and repeated stress responses.

These findings contribute to our understanding of how early developmental environments shape adult stress phenotypes. They also highlight potential mechanisms underlying behavioural plasticity in invasive species, where adaptability to repeated stressors may facilitate persistence in novel environments. Future work should aim to link behavioural and physiological responses across ontogeny and explore neural mechanisms that underpin different forms of learning and habituation.

#### Cold incubation temperature increases the response to an acute stressor in _L. delicata_ but not in _L. guichenoti_
**CHECK PROPER UNITS**
Hot-incubated _L. delicata_ spent less time frozen after the simulated attack. We did not see any differnces betwen treatments in any other variable recorded. No treatment effect on the response to an acute stressor in _L. guichenoti_.
- Why do we consider less time to be less reactive?
- Why differences in temp? --> Cold animals had higher baseline CORT levels...but CORT treated too...at least mean treatment but not high.
- Why not differences based on CORT?
- Why is temperature in _L. delicata_ more important for this than in _L. guichenoti_?? Invassiveness??
